Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-12 talk/pepm2011/paper.tex
add missing citations

The most important technique we use to achieve to optimize traces is a
straightforward application of partial evaluation. The partial evaluation
performs a form of escape analysis \cite{XXX} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}, a more formal description is
given in Section~\ref{sec:formal}.

@@

Tracing JITs are a recently popular approach to write just-in-time compilers for
dynamic languages. Their origins lie in the Dynamo project, which used a tracing
approach to optimize machine code using execution traces
\cite{bala_dynamo:_2000}. Tracing JITs have then be adapted to be used for a
very light-weight Java VM \cite{gal_hotpathvm:_2006} and afterwards used in
several implementations of dynamic languages, such as JavaScript
\cite{andreas_gal_trace-based_2009}, Lua \cite{XXX} and now Python (and other
languages) via PyPy.

@@

This problem is solved by making it possible to the interpreter author to add
some hints into the source code to declare instances of one class as frame
objects. The JIT will then fill these objects only lazily when they are actually
accessed (\eg because a debugger is used). Therefore in the common case, nothing
is stored into the frame objects, making the problem of too much escaping go
away. This is a common approach in VM implementations \cite{XXX}, the only
novelty in our approach lays in its generality, because most other JITs are just
specifically written for one particular language.

@@

Furthermore there are the following benchmarks:
\begin{itemize}
    \item \textbf{crypto\_pyaes}: AES implementation.
    \item \textbf{django}: The templating engine of the Django web
    framework\footnote{\texttt{http://www.djangoproject.com/}}.
    \item \textbf{go}: A Monte-Carlo Go
    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
    \item \textbf{html5lib}: HTML5 parser
    \item \textbf{pyflate-fast}: BZ2 decoder
    \item \textbf{raytrace-simple}: ray tracer
    \item \textbf{richards}: The Richards benchmark \cite{XXX}
    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
    \item \textbf{telco}: A Python version of the Telco decimal
    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
    using a pure Python decimal floating point implementation.
    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
    framework\footnote{\texttt{http://twistedmatrix.com/}}.
\end{itemize}

@@

In addition to the count of operations we also performed time measurements. The
machine the benchmarks were run on is XXX. We compared the performance of
various Python implementations on the benchmarks. As a baseline, we used the
standard Python implementation in C, called
CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against Psyco \cite{XXX}, an extension to
CPython which is a just-in-time compiler that produces machine code at run-time.
It is not based on traces. Of PyPy's Python interpreter we used three versions,
one without a JIT, one including the JIT but not using the allocation removal
optimization, and one using the allocation removal optimizations.

All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{XXX}. The results are reported in Figure~\ref{fig:times}.
With the optimization turned on, PyPy's Python interpreter outperforms CPython
in all benchmarks except spambayes (which heavily relies on regular expression
performance). All benchmarks are improved by the allocation removal
optimization, some as much as XXX. XXX Psyco

@@

SPUR, a tracing JIT for C\# seems to be able to remove allocations in a similar
way to the approach described here, as hinted at in the technical report
\cite{XXX}. However, no details for the approach and its implementation are
given.

>>>>

The most important technique we use to achieve to optimize traces is a
straightforward application of partial evaluation. The partial evaluation
performs a form of escape analysis \cite{bruno_blanchet_escape_2003} on the traces and make some
objects that are allocated in the trace \emph{static}\footnote{These objects are
called \emph{virtual} in Psyco \cite{rigo_representation-based_2004}.} which
means that they do not occur any more in the optimized trace. This technique is
informally described in Section~\ref{sec:statics}, a more formal description is
given in Section~\ref{sec:formal}.

@@

Tracing JITs are a recently popular approach to write just-in-time compilers for
dynamic languages. Their origins lie in the Dynamo project, which used a tracing
approach to optimize machine code using execution traces
\cite{bala_dynamo:_2000}. Tracing JITs have then be adapted to be used for a
very light-weight Java VM \cite{gal_hotpathvm:_2006} and afterwards used in
several implementations of dynamic languages, such as JavaScript
\cite{andreas_gal_trace-based_2009}, Lua\footnote{\texttt{http://luajit.org/}}
and now Python (and other languages) via PyPy.

@@

This problem is solved by making it possible to the interpreter author to add
some hints into the source code to declare instances of one class as frame
objects. The JIT will then fill these objects only lazily when they are actually
accessed (\eg because a debugger is used). Therefore in the common case, nothing
is stored into the frame objects, making the problem of too much escaping go
away. This is a common approach in VM implementations \cite{miranda_context_1999}, the only
novelty in our approach lays in its generality, because most other JITs are just
specifically written for one particular language.

@@

Furthermore there are the following benchmarks:
\begin{itemize}
    \item \textbf{crypto\_pyaes}: AES implementation.
    \item \textbf{django}: The templating engine of the Django web
    framework\footnote{\texttt{http://www.djangoproject.com/}}.
    \item \textbf{go}: A Monte-Carlo Go
    AI\footnote{\texttt{http://shed-skin.blogspot.com/2009/07/ disco-elegant-python-go-player.html}}.
    \item \textbf{html5lib}: An HTML5 parser.
    \item \textbf{pyflate-fast}: A BZ2 decoder.
    \item \textbf{raytrace-simple}: A ray tracer.
    \item \textbf{richards}: The Richards benchmark.
    \item \textbf{spambayes}: A Bayesian spam filter\footnote{\texttt{http://spambayes.sourceforge.net/}}.
    \item \textbf{telco}: A Python version of the Telco decimal.
    benchmark\footnote{\texttt{http://speleotrove.com/decimal/telco.html}},
    using a pure Python decimal floating point implementation.
    \item \textbf{twisted\_names}: A DNS server benchmark using the Twisted networking
    framework\footnote{\texttt{http://twistedmatrix.com/}}.
\end{itemize}

@@

In addition to the count of operations we also performed time measurements. The
machine the benchmarks were run on is XXX. We compared the performance of
various Python implementations on the benchmarks. As a baseline, we used the
standard Python implementation in C, called
CPython\footnote{\texttt{http://python.org}}, which uses a bytecode-based
interpreter. Furthermore we compared against Psyco \cite{rigo_representation-based_2004}, an extension to
CPython which is a just-in-time compiler that produces machine code at run-time.
It is not based on traces. Of PyPy's Python interpreter we used three versions,
one without a JIT, one including the JIT but not using the allocation removal
optimization, and one using the allocation removal optimizations.

All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in Figure~\ref{fig:times}.
With the optimization turned on, PyPy's Python interpreter outperforms CPython
in all benchmarks except spambayes (which heavily relies on regular expression
performance). All benchmarks are improved by the allocation removal
optimization, some as much as XXX. XXX Psyco

@@

SPUR, a tracing JIT for C\# seems to be able to remove allocations in a similar
way to the approach described here, as hinted at in the technical report
\cite{michael_bebenita_spur:_2010}. However, no details for the approach and its
implementation are given.
